Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-05 tagless/tagless-final.tex
almost done the related work.

\jacques{Got to here}.
Regarding our implementation of the type-level function |'sv -> 'dv|.
Gibbons et al paper on
the implementations of typecase may be relevant. I have not read it,
but I have the feeling some of their typecase designs may be
relevant (at least to cite). 
However, the typecase paper relies on having a 
bijection between types -- our |'sv| and |'dv| in this case, which is not 
our situation.  We have a one-sided view of things, where |'sv -> 'dv| is 
easy, and |'dv| "perfectly reflects" an |'sv|, but you still can't get at 
that |'sv|.  \jacques{Could we say that we have a meta-level Galois
injection?  We have one direction completely covered at the object
level (and so easy to lift), but 'dv to 'sv can only be done by
an outside mechanism}

Self-Interpretation and Reflection in a Statically Typed Language 
\url{http://webpages.cs.luc.edu/~laufer/papers/reflection93.pdf}
He uses the SK language and some typeclasses. Their interpreter is a
meta-circular interpreter -- rather than self-interpreter. We can
claim to have implemented a far better interpreter for a far richer
language.

co-algebra references.  Many.

>>>>

When dealing with datatypes which are equivalent but have different
representations, it is customary to use injection-projection pairs
(used in the context of interpreters by \citet{Ramsey-ML-module-mania}
and \citet{Benton-embedded-interpreters}, partial evaluation by
\citet{Danvy-TDPE}, and type-level functions by
\citet{oliveira-typecase}).  Na\"{\i}vely, this would appear to apply to
our implementation of the type-level function |'sv -> 'dv|.  However, in most
of these cases, it is important that this establishes a bijection (between
|'sv| and |'dv| in our case).  Our approach is one-sided: we requite
an operational definition of |'sv -> 'dv| (which is easy), and an
implementation of |'dv| which ``perfectly reflects'' |'sv| without 
however making this uniformly available.

\citet{fiore:nbe-ppdp2002,balat:tdpe-popl2004} also deal with a 
tagfree partial-evaluator using quite a lot of interesting 
technology (amongst them powerful continuation-based control operators).
While they use a reify-reflect injection-projection pair, this appears
to be done ``lightly'', in the sense that the structure of compiled code
is never examined.  The exact relation between their work and ours is 
still a little mysterious to us, but we will continue to investigate it.

We are not the first to implement a typed interpreter for a typed
language -- \citet{laod93} use a typed version of the SK language
(and typeclasses) to implement a meta-circular interpreter (rather than
a self-interpreter).  Interestingly, their meta-circular interpreter
appears to be tagfree!  However, they could not have implemented a
compiler or a partial-evaluator in the same way since they rely quite
heavily on having injection-projection pairs (through reflection and
reification).

We have been unable to find a direct reference that establishes that
the typed $\lambda$-calculus possesses a final co-algebra structure.
For the untyped $\lambda$-calculus, 
\citet{HonsellLenisa,honsell99coinductive} establish several interesting
results along this line.  Of particular interest is their use of
contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
of \emph{observational equivalence}, which directly relates to our
use of terms with a hole as a context for our self-interpreter 
(\S\ref{selfinterp}).  The reader is directed to bibliography of
\citet{honsell99coinductive} for references to the foundational work
in this important area.

\jacques{Dig through the last couple of emails I just got from the
category theory list with references, and this section will be done.}
